Making multiple-locus association mapping on a genome-wide scale routine

### 2017-11-09 AM+draft.tex
10Nov17 Morning commit

BigRR is an R package that implements generalized ridge regression. Generalized ridge regression is 
 an improvement on ridge regression. In ridge regression, coefficients of the predictors are shrunk 
to zero by a constant amount.  In generalized ridgre regression, the amount of shrinkage is allowed to vary, giving opportunity for 
a more realistic model to be specified. The model used in BigRR is 
$y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of snp effects, and each snp 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$.

BigRR is an R package that implements generalized ridge regression
which is an improvement on ridge regression. In ridge regression, coefficients of the predictors are shrunk 
to zero by a constant amount.  In generalized ridgre regression, the amount of shrinkage is allowed to vary, giving opportunity for 
a more realistic model to be specified. The model used in BigRR is 
$y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of snp effects, and each snp 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$.

BigRR, LMM-Lasso, and glmnet all implement regularization methods for parameter
estimation. Regularization methods are well suited to problems where the number of predictors is far greater than the 
number of samples. It is not surprising then that regularization methods have attracted considerable interest, primarily for 
genomic prediction (refs) but also association mapping (refs). 
BigRR implements generalized ridge regression, LMM-Lasso implements a two-step lasso approach, and glmnet makes use of 
elastic net for parameter estimation.

BigRR \cite{shen2013novel}  is an R package that implements generalized ridge regression.
In ridge regression, the coefficients of the predictors are shrunk 
to zero by a constant amount.  In generalized ridge regression, the amount of shrinkage is allowed to vary, giving opportunity for 
a more realistic model to be specified. The model used in BigRR can be formulated as a linear mixed model that is 
$y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of snp effects, and each snp 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$. 
The parameters are estimated via a 
algorithm that makes clever use of the Cholesky decomposition of the genomic relationship matrix, 
and the equivalence between the above model and animal models \ref{stranden2009animalmodel}. 
One of the challenges in using BigRR for association mapping is that while it has the capacity to estimate the effect sizes of all the snp simultaneously, 
obtaining their statistical significance is difficult. 
Permutation has been suggested as a way of calculating significance 
empirically \cite{shen2013novel} but we instead opted for stability selection.

Lmm-Lasso  \cite{rakitsch2013lasso} is a stand-alone computer program that implements lasso within a linear mixed model 
framework. Unlike in ridge regression, lasso performs parameter estimation and variable selection, simultaneously.   The 
coefficients of "unimportant" predictors are shrunk to exactly zero.
The (full) model is 
$y =  \sum_{l=1}^L m_l \theta_l +   g  + e$ where $g^{(n ; \times 1)}$ is a vector of random polygenic effects such that 
$g \sim N(0, \sigma^2_g K)$, and $K^{(n ; \times \; n)}$ is a realised relationship matrix reconstructed from the snp information. 
It is challenging, computationally,  to apply lasso directly to the full model for parameter estimation. Instead, a two step procedure 
is adopted. In the first step, the above model without any snp effects is fitted. In the second step, 
the unknown variance components in the full model are replaced by their estimates 
that were calculated in the first step. Lasso is then applied to this simpler model making lasso more tractable, computationally. 
As with BigRR,  the statistical significance of the snp effects are not calculated. We use stability selection to obtain the 
significance empirically.

Glmnet  \cite{Friedman2010glmnet}, unlike BigRR and LMM-Lasso, was not purpose built for 
association mapping.  It is a general R package for 
regression analysis. We have included it in our list of applications for comparison with Eagle because the 
regularization method, elastic net, upon which the package is based has been found to be superior to ridge regression and 
lasso (ref needed???).  Elastic net uses a penalty function that is a mixture of 
$\ell_1$-norm (lasso) and $\ell_2$-norm (ridge regression) penalties  \cite{zou2005regularization}. 
This allows elastic net to avoid the limitations of lasso 
while still being able to perform shrinkage and variable selection, simultaneously. 
Glmnet can handle a range of different models. For the analyses performed in this paper, the model is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the coefficient for the $l$th snp. 
We used stability selection to obtain the significance of the coefficients (snp effects).

Of the multiple-locus applications considered in this paper, 
MLMM \cite{segura2012efficient} bares the greatest similarity to Eagle algorithmically.  MLMM, like Eagle, performs 
association mapping by building the "best" linear mixed model.  The model used for analysis 
is the same as (\ref{eq1}) except that the genetic effect is 
\begin{equation}
\label{eqmlmm}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + g
\end{equation}
where 
$g^{(n ; \times 1)}$ is a vector of random polygenic effects with distribution 
$g \sim N(0, \sigma^2_g K)$. The model is built iteratively, via forward-backward variable selection.  In Eagle, at each iteration, 
the snp in strongest association with the trait is found by calculating the score statistic $t_j^2$.  In MLMM, at each iteration, 
a separate linear mixed model analysis is performed for each snp not already selected. The snp yielding the most 
significant result (lowest p-value) is selected for inclusion in the fixed effects part of (\ref{eqmlmm}).  
Two selection 
criteria are available for the model building process: 
the EBIC and the multiple-Bonferroni criterion. We used only the EBIC.

@@

@@

\textbf{gemma}

Fast-LMM

>>>>

\textbf{BigRR:} BigRR \cite{shen2013novel}  is an R package that implements generalized ridge regression.
In ridge regression, the coefficients of the predictors are shrunk 
to zero by a constant amount.  In generalized ridge regression, the amount of shrinkage is allowed to vary, giving opportunity for 
a more realistic model to be specified. The model used in BigRR can be formulated as a linear mixed model that is 
$y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of snp effects, and each snp 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$. 
The parameters are estimated via a 
algorithm that makes clever use of the Cholesky decomposition of the genomic relationship matrix, 
and the equivalence between the above model and animal models \ref{stranden2009animalmodel}. 
One of the challenges in using BigRR for association mapping is that while it has the capacity to estimate the effect sizes of all the snp simultaneously, 
obtaining their statistical significance is difficult. 
Permutation has been suggested as a way of calculating significance 
empirically \cite{shen2013novel} but we instead opted for stability selection.

\textbf{LMM-Lasso:} LMM-Lasso  \cite{rakitsch2013lasso} is a stand-alone computer program that implements lasso within a linear mixed model 
framework. Unlike in ridge regression, lasso performs parameter estimation and variable selection, simultaneously.   The 
coefficients of "unimportant" predictors are shrunk to exactly zero.
The (full) model is 
$y =  \sum_{l=1}^L m_l \theta_l +   g  + e$ where $g^{(n ; \times 1)}$ is a vector of random polygenic effects such that 
$g \sim N(0, \sigma^2_g K)$, and $K^{(n ; \times \; n)}$ is a realised relationship matrix reconstructed from the snp information. 
It is challenging, computationally,  to apply lasso directly to the full model for parameter estimation. Instead, a two step procedure 
is adopted. In the first step, the above model without any snp effects is fitted. In the second step, 
the unknown variance components in the full model are replaced by their estimates 
that were calculated in the first step. Lasso is then applied to this simpler model making lasso more tractable, computationally. 
As with BigRR,  the statistical significance of the snp effects are not calculated. We use stability selection to obtain the 
significance empirically.

\textbf{Glmnet:} Glmnet  \cite{Friedman2010glmnet}, unlike BigRR and LMM-Lasso, was not purpose built for 
association mapping.  It is a general R package for 
regression analysis. We have included it in our list of applications for comparison with Eagle because the 
regularization method, elastic net, upon which the package is based has been found to be superior to ridge regression and 
lasso (ref needed???).  Elastic net uses a penalty function that is a mixture of 
$\ell_1$-norm (lasso) and $\ell_2$-norm (ridge regression) penalties  \cite{zou2005regularization}. 
This allows elastic net to avoid the limitations of lasso 
while still being able to perform shrinkage and variable selection, simultaneously. 
Glmnet can handle a range of different models. For the analyses performed in this paper, the model is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the coefficient for the $l$th snp. 
We used stability selection to obtain the significance of the coefficients (snp effects).

\textbf{MLMM:} Of the multiple-locus applications considered in this paper, 
MLMM \cite{segura2012efficient} bares the greatest similarity to Eagle algorithmically.  MLMM, like Eagle, performs 
association mapping by building the "best" linear mixed model.  The model used for analysis 
is the same as (\ref{eq1}) except that the genetic effect is 
\begin{equation}
\label{eqmlmm}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + g
\end{equation}
where 
$g^{(n ; \times 1)}$ is a vector of random polygenic effects with distribution 
$g \sim N(0, \sigma^2_g K)$. The model is built iteratively, via forward-backward variable selection.  In Eagle, at each iteration, 
the snp in strongest association with the trait is found by calculating the score statistic $t_j^2$.  In MLMM, at each iteration, 
a separate linear mixed model analysis is performed for each snp not already selected. The snp yielding the most 
significant result (lowest p-value) is selected for inclusion in the fixed effects part of (\ref{eqmlmm}).  
Two selection 
criteria are available for the model building process: 
the EBIC and the multiple-Bonferroni criterion. We used only the EBIC.

@@

We were also interested in comparing Eagle to single locus methods to measure the difference in run times and statistical power. Even by limiting our focus to LMM-based methods, there were a number of efficient applications to choose from. 
We decided on GEMMA and FaST-LMM \cite{lippert2011fast}. These two applications have the same computational complexity (ref), produce exact instead of approximate results, and are highly efficient computationally. They were developed at the same time, independently, but are very similar theoretically. Both perform a single spectral decomposition of the relationship matrix K. Both use the eigenvector matrix to rotate the data. Both reformulate the log likelihood and reml log likelihood into a sum of $n$ terms that are easier to compute. The difference lies in their estimation procedure. FaST-LMM implement's the Brent algorithm to optimise $\delta$. GEMMA instead implement's the Newton Raphson algorithm. Newton Raphson is more complicated in that it also requires the first and second derivatives of a function to be calculated. However, it is superior in terms of its convergence properties to the Brent algorithm (quadratic verse super-linear).  
Both applications are stand along computer programs, popular, and in common use.

\section{Stability Selection}

\section{Simulation Study}

\section{Implementation}

@@
